---
title: "CHIC402/CHIC602 Coursework 2 - Variant Analysis"
author: 36 ...
output: 
  pdf_document:
    number_sections: true
---


<!--- Global Settings --->
```{r include = FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

<!--- Libraries --->
```{r include = FALSE}
library(dplyr)
library(ggplot2)
```

<!--- External Functions --->
```{r include = FALSE}
# Uploads functions DataGet & DataPreview
sys.source(file = 'R/functions/VariantsData.R', envir = knitr::knit_global())

sys.source(file = 'R/functions/GraphsDelta.R', envir = knitr::knit_global())
```


This project focuses on the analysis of the United Kingdom's genetically-typed SARS-CoV-2 infections data.  The data set summarises the weekly infections proportions of $17$ SARS-CoV-2 variants, starting from $31$ October 2021; the field `Other` consists of the pooled fractions of an unknown number of unnamed variants.


## The Variants Data

```{r include = FALSE}
variants <- DataGet()
preview <- DataPreview(variants = variants)
```

\vspace{20pt}

```{r size = 'tiny', comment = '    '}
str(variants, vec.len = 2)
```

\vspace{10pt}

```{r  size = 'tiny', comment = '', results = 'asis'}
# paste(preview$variant, collapse = ', ')
```


## An Exploration of Non-travel Data

Each variant's weekly data consists of two records.  One represents the infections-fraction associated with known travel links (denoted *travel* in field `travel`), whilst the other represents the infections-fraction without known travel links (denoted *none* in field `travel`).

```{r}
# 3.0
# The non-travel data only
local <- variants %>%
        dplyr::filter(travel == 'none')
```


```{r out.width = '65%', fig.align = 'center'}
# 3.1.a
# Delta proportions over time
GraphsDelta(local = local)
```






